---
layout: post
title: "Monte Carlo of Random Correlations"
categories: blog
author: "Keith Hultman"
date: 2016-10-29
tags: [BigData, correlation]
output: html_document
---

# Exploring correlations of random numbers

When working with big data you need to be more aware of statistical outliers than you do with more typical data sizes. Basic statistical tests like a Student's t-test or Pearson correlation are acceptable when you only test a few relationships in a small data set. But when you examine the correlation with thousands of columns of data, you are bound to find several that are strongly correlated. It's always good to have an idea for the 'expect' value. How many correlations should I expect to see with this R value or better given the size of my data set.

To explore this idea, we can take a look at a random set of features produced by [Random.org](http://www.random.org). These data are organized into a 1000x20 table (1000 rows and 20 columns). We can create a correlation matrix and view the plot of Pearson's correlation for each pairing of the 20 columns. 

```{r}
setwd("~/datasets/")
randata <- read.csv("Random\ data\ v2.csv", header = FALSE)
library(corrplot)
randcorr <- cor(randata)
corrplot(randcorr, method = "circle")
```

With 20 features and 1000 observations for each feature we don't see many features that have a high correlation. The highest correlation value, in either positive or negative direction is:

```{r}
randcorr[randcorr == 1] <- 0
max(abs(randcorr))
```
The correlation of 0.104 is a very low value, suggesting that there little to no relationship between our features. What happens when we transpose the data frame to have 1000 features with only 20 observations? Let's look for strong correlations, with an R > 0.8.

```{r}
trandata <- t(randata)
trandcorr <- cor(trandata)
trandcorr[trandcorr == 1] <- 0 
high_corr <- trandcorr > 0.8
set.seed(5)
library(caret)
highlyCorrelated <- findCorrelation(trandcorr, cutoff = 0.8, verbose = TRUE)
length(highlyCorrelated)
```
There are 14 pairs of columns that have a Pearson's correlation above 0.8 in this data set. 

I am not sure how to extract all of the highly correlated columns programatically, the findCorrelation function will only report one of the two highly correlated columns since it is used to remove columns in a dataset that are correlated and therefore redundant. In this case I want to extract just those pairs of columns, so I create the list of missing columns by hand below. Then we can plot the correlations of the columns with correlations above 0.8.

```{r}
hc <- c(177,407,48,266,178,439,65,92,380,589,759,134,48,197,80)
highlyCorrelated <- c(highlyCorrelated, hc)
trandata_high <- cor(trandata[,highlyCorrelated])
corrplot(trandata_high, method = "circle")
```

### And knowing is half the battle!
Here we can see several strong correlations between these columns. Strong positive correlations are dark blue and strong negative correlations are dark red. This Monte Carlo simulation shows how easy it is to find very strong correlations in a wide and shallow data set even from completely random numbers. 